---
title: "Clinical Trial Simulation Report"
author: "Suprise Baloyi"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(ggplot2)
library(gt)
library(knitr)
set.seed(123)
```

# Introduction

This report presents a **simulated Phase II clinical trial** comparing a new treatment (**DrugA**) to **Placebo**.  
The goal is to demonstrate basic statistical programming and reporting principles using R — from data simulation to summary tables, visualization, and inferential analysis.

# 1. Study Population Simulation

```{r simulate}
n <- 60
data <- data.frame(
  USUBJID = sprintf("SUBJ%03d", 1:n),
  TRT01A = rep(c("Placebo", "DrugA"), each = n/2),
  AGE = round(rnorm(n, 50, 8)),
  SEX = sample(c("M", "F"), n, replace = TRUE)
)
head(data)
```

# 2. Baseline and Week 4 Measurements

```{r baseline}
data <- data %>%
  mutate(
    BASE = round(rnorm(n, 140, 10), 1),
    WEEK4 = ifelse(TRT01A == "DrugA",
                   BASE - rnorm(n, mean = 10, sd = 5),
                   BASE - rnorm(n, mean = 3, sd = 5)),
    CHG = WEEK4 - BASE,
    AE_FLAG = sample(c(0, 1), n, replace = TRUE, prob = c(0.7, 0.3))
  )
head(data)
```

# 3. ADaM-like Dataset

```{r adam}
adam <- data %>% select(USUBJID, TRT01A, AGE, SEX, BASE, WEEK4, CHG, AE_FLAG)
head(adam)
```

# 4. Table 1 – Baseline Demographics

```{r table1}
table1 <- adam %>%
  group_by(TRT01A) %>%
  summarise(
    N = n(),
    Mean_Age = round(mean(AGE), 1),
    SD_Age = round(sd(AGE), 1),
    Female = sum(SEX == "F"),
    Male = sum(SEX == "M")
  )

gt(table1) %>% tab_header(title = "Table 1: Baseline Demographics")
```

# 5. Table 2 – Efficacy (Change from Baseline)

```{r efficacy}
efficacy <- adam %>%
  group_by(TRT01A) %>%
  summarise(
    Mean_BASE = round(mean(BASE), 1),
    Mean_WEEK4 = round(mean(WEEK4), 1),
    Mean_CHG = round(mean(CHG), 1),
    SD_CHG = round(sd(CHG), 1)
  )

t_test <- t.test(CHG ~ TRT01A, data = adam)
p_value <- round(t_test$p.value, 3)
efficacy$P_value <- c("", p_value)

gt(efficacy) %>% tab_header(title = "Table 2: Efficacy Results – Change in Systolic BP")
```

# 6. Table 3 – Adverse Events

```{r ae}
ae_summary <- adam %>%
  group_by(TRT01A) %>%
  summarise(
    N = n(),
    AE_Count = sum(AE_FLAG),
    AE_Percent = round(mean(AE_FLAG) * 100, 1)
  )

gt(ae_summary) %>% tab_header(title = "Table 3: Adverse Events Summary")
```

# 7. Visualization

## 7.1 Change in Systolic BP by Treatment Group

```{r boxplot}
ggplot(adam, aes(x = TRT01A, y = CHG, fill = TRT01A)) +
  geom_boxplot() +
  labs(
    title = "Change in Systolic Blood Pressure (Week 4 – Baseline)",
    x = "Treatment Group",
    y = "Change in SBP (mmHg)"
  ) +
  theme_minimal()
```

## 7.2 Mean SBP Over Time by Treatment

```{r trend}
adam_long <- adam %>% tidyr::pivot_longer(cols = c(BASE, WEEK4), names_to = "Visit", values_to = "SBP")

ggplot(adam_long, aes(x = Visit, y = SBP, group = USUBJID, color = TRT01A)) +
  geom_line(alpha = 0.4) +
  stat_summary(fun = mean, geom = "line", size = 1.5, aes(group = TRT01A)) +
  labs(title = "Mean Systolic BP Over Time by Treatment", y = "Systolic BP (mmHg)") +
  theme_minimal()
```

# 8. Output Files

```{r save}
write.csv(adam, "adam_dataset.csv", row.names = FALSE)
write.csv(table1, "table1_baseline.csv", row.names = FALSE)
write.csv(efficacy, "table2_efficacy.csv", row.names = FALSE)
write.csv(ae_summary, "table3_ae.csv", row.names = FALSE)
```

# Conclusion

This simulation illustrates how a clinical trial analysis pipeline can be implemented in R — covering data generation, descriptive statistics, hypothesis testing, and graphical reporting.  
Such projects strengthen statistical programming and reporting skills, relevant for roles in **clinical data science, biostatistics, and statistical programming**.
